Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-07 numerical.tex
merge with youdin's updates
new 6.4, merge with youdin's updates

\begin{figure}
  %\includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  %0.9cm]{figures/gcorr_compare_iso_maxrate_z3} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z5} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z7}  
  \caption{Maximum VSI growth rates in the fiducial disk 
     model as a function of the thermal relaxation timescale
     $\beta$, for %$\zmax=3H$ (top), 
     $\zmax=5H$ (top) and $\zmax=7H$ (bottom). The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

For $\khat = 5, 10$, the growth rate drops to zero at the expected $\beta = \beta_\mathrm{crit}$.
For longer wavelength modes, with $\khat = 1$, growth persists to slightly longer cooling times.  
This difference is not surprising since our analytic derivation assumed $\khat \gg1$, and 
the change is quantitatively minor.

@@

\subsection{Vertically non-isothermal disk model} 
%\emph{fix this rambling discussion..}

% \citetalias{barker15} have argued that vertically isothermal disks are not
% ideal for studing the VSI because $\zmax$ is a free 
% parameter in numerical calcluations. On the other hand, vertically 
% non-isothermal disks have a disk surface at finite height so there is
% a natural choice for $\zmax$. This means that higher order body modes and
% surface modes are better defined in a vertically non-isothermal disk.

% However, as we have seen in Fig. \ref{bcrit_compare1}, the maximum
% growth rate becomes insensitive to $\zmax$ for $\beta\gtrsim
% \beta_\mathrm{crit}$. This suggests that the characteristic upper
% limit on the thermal timescale for the VSI is not sensitive to
% $\zmax$, so that vertically isothermal disks are in fact suitable for
% the purpose of characterizing the thermodynamic dependence of the
% VSI.

%Physically this is because modes that depend on vertical boundaries
%are also subject to the strongest stabilization by buoyancy, which
%maximize at $\zmax$. These modes are theremore always strongly damped
%as we increase the thermal timescale.

Here we briefly consider a vertically non-isothermal
disk model with $(p,q, h)=(0,-1,0.05)$ and $\Gamma=1.0837$. This
value of $\Gamma$ is chosen so that the standard vertical domain size $\zmax =
5H\simeq H_s$ effectively corresponds to the disk surface. The disk
temperature decreases rapidly with height such that
$c_s^2(\zmax)\simeq 10^{-4}c_s^2(0)$. A disk surface at finite height
means that higher order body modes and surface modes are better
defined (cf. in vertically isothermal disk models where the disk
boundary $\zmax$ is a free parameter, see \citetalias{barker15}).

% \emph{note that this would
% imply midplane 
% temperatures in excess of $10^6$ K so it's completely un-astrophyscial.}

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=10$ for several values of $\beta$. This plot is 
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order body modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely suppressed.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/compare_modes_Gam1.084_kx10.ps}
  \caption{Unstable modes in a vertically non-isothermal disk with
    $\khat=10$ and a range of thermal relaxation timescales.  
    \label{compare_modes_vnoniso_kx10}}
\end{figure}

We do not find surface modes for $\khat=10$, but do find them for
sufficiently large $\khat$. We demonstrate this in  % In the vertically
% non-isothermal disk, surface modes develop for sufficiently large 
% $\khat$ \citep{barker15}. We demonstrate them in
Fig. \ref{compare_modes_vnoniso_kx100} which show eigenvalues for
$\khat=100$. While growth rates generally decrease with increasing 
$\beta$, we find eigenvalues can overlap, making it difficult to 
distinguish between different modes. However, since there 
is a well-defined surface, we may seek the  
most unstable mode without the boundary-dependent issue that arose 
with a vertically isothermal disk in the limit $\beta\to0$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/compare_modes_Gam1.084_kx100.ps}
  \caption{Same as Fig. \ref{compare_modes_vnoniso_kx10} but with
    $\khat=100$. In this case surface modes appear nearly parallel to
    the vertical axis
    (cf. Fig. \ref{compare_modes_vnoniso_kx10}). However, this
    distinction becomes difficult for increasing $\beta$ as modes
    overlap. 
    % but
    % overlap with body modes (those that are roughly horizontal when 
    % $\beta=10^{-3}$) as $\beta$ is further increased. 
    \label{compare_modes_vnoniso_kx100}}
\end{figure}

Fig. \ref{compare_eigenvz_vnoniso_kx100} show the most unstable modes
with $\khat=100$ and $\beta=10^{-3},\, 10^{-2}$ and
$0.1$. For $\beta=10^{-3}$ the most 
unstable mode is an asymmetric surface mode \citepalias[also seen in Fig. 7
of][]{barker15}. However, increasing the thermal timescale to
$\beta=10^{-2}$ an anti-symmetric surface mode becomes
dominant. Further increasing to $\beta=0.1$, the dominant mode has a
mixed character between a surface mode and the fundamental
mode. Notice as $\beta$ is increased the perturbation amplitude near
the vertical boundaries become negligible. This is not surprsing since  
the stabilizing effect of buoyancy is largest at the disk surface. 
%formally diverges

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.73cm 0cm
  0cm]{figures/eigenvectorvz_vnoniso_kx100_beta0d001.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.73cm 0cm
  0cm]{figures/eigenvectorvz_vnoniso_kx100_beta0d01.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorvz_vnoniso_kx100_beta0d1.ps}
  \caption{Vertical velocity eigenfunction of the most unstable mode
    with $\khat=100$ found in the vertically non-isothermal disk with
    $\beta = 10^{-3}$ (top), $\beta=10^{-2}$ (middle) and $\beta=0.1$
    (bottom). 
    \label{compare_eigenvz_vnoniso_kx100}}
\end{figure}

Finally, in Fig. \ref{bcrit_compare2} we plot the maximum VSI growth
rates in the vertically non-isothermal disk model. Based on 
Eq. \ref{iso_vsi_cond}, we conjecture the critical 
thermal timescale in the vertically non-isothermal disk is 
given by $ h|s|/(\gamma - \Gamma)$. This 
evaluates to 0.158, which is close to the cut-off thermal timescale
for $\khat\in[5,30]$. Fig. \ref{bcrit_compare2} is qualitatively 
similar to the fiducial disk model in
Fig. \ref{bcrit_compare1}. Importantly, for $\khat=O(10)$ there is a
precise cut-off thermal timescale, the value of which in fact provides a  
characteristic upper limit for other $\khat$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gcorr_compare_vnoniso_maxrate} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $\beta$. The vertical dashed line is the critical thermal
    timescale given by Eq. \ref{iso_vsi_cond}. A conjectured critical thermal
    timescale accounting for a non-isothermal background, given by 
    $\beta_\mathrm{crit,noniso}= h|s|/(\gamma-\Gamma)=0.158$,  
    is also shown as the vertical dotted line. 
    \label{bcrit_compare2}}   
\end{figure}

@@

>>>>

\begin{figure}
  %\includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  %0.9cm]{figures/gcorr_compare_iso_maxrate_z3} 
  \includegraphics[scale=0.4415,clip=true,trim=0cm 1.75cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z5} 
  \includegraphics[scale=0.4415,clip=true,trim=0cm 0.0cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z7}  
  \caption{Maximum VSI growth rates in the fiducial disk 
     model as a function of the thermal relaxation timescale
     $\beta$, for %$\zmax=3H$ (top), 
     $\zmax=5H$ (top) and $\zmax=7H$ (bottom). The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

For $\khat = 5, 10$, the growth rate drops to zero at the expected $\beta = \beta_\mathrm{crit}$.
For longer wavelength modes, with $\khat = 1$, growth persists to slightly longer cooling times.  
This difference is not surprising since our analytic derivation assumed $\khat^2 \gg1$, and 
the change is quantitatively minor.

@@

\subsection{Vertically non-isothermal disk model} 
We briefly consider a vertically non-isothermal disk model with
$(p,q, h)=(-0.5,-1,0.05)$ and $(\gamma, \Gamma)=(1.4,1.3)$. We set the
vertical domain size close to the physical disk surface, $\zmax =
0.99H_s\simeq 2.6H$, which is smaller than our standard domain
size. In this case the vertical shear rate $\p_z\Omega\propto s$, the
radial entropy gradient. Here, $s=-0.85$.

From the expression for the
critical cooling timescale, Eq. \ref{iso_vsi_cond}, it is reasonable
to expect a similar criteria for vertically non-isothermal 
disks in the form
\begin{align}\label{bcrit_noniso}
 \beta_\mathrm{crit}\to\beta_\mathrm{crit, noniso} =
 \frac{h|s|}{\gamma - \Gamma},
\end{align}
and $\beta_\mathrm{crit,noniso}=0.425$ for the chosen parameters. This
is $\sim 3$ times larger than our fiducial, vertically isothermal
disk with $\beta_\mathrm{crit}=0.125$.

% A disk surface at finite height
% means that higher order body modes and surface modes are better
% defined (cf. in vertically isothermal disk models where the disk
% boundary $\zmax$ is a free parameter, see \citetalias{barker15}).

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=30$ for several values of $\beta$. This plot is 
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order body modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely suppressed. (Note the significant reduction in
growth rates for $\beta=0.4$, which is close to
$\beta_\mathrm{crit,noniso}$.) However, this behavior is not seen for 
large wavenumbers (see below).

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/compare_modes_Gam1.3_kx30.ps}
  \caption{Unstable modes in a vertically non-isothermal disk with
    $\khat=30$ and a range of thermal relaxation timescales.  
    \label{compare_modes_vnoniso_kx10}}
\end{figure}

In Fig. \ref{bcrit_compare2} we plot the maximum VSI growth
rates in the vertically non-isothermal disk model, which is
qualitatively similar to Fig. \ref{bcrit_compare1} for the vertically
isothermal disk. For $\khat=O(10)$, there is a   
precise cut-off thermal timescale close to that given by
Eq. \ref{bcrit_noniso}. We found that very large wavenumbers, 
$\khat > O(10^2)$, were needed to enable surface modes, and modes with
such values of $\khat$ do not have a maximum cooling time. However,
these small lengthscales may not be important in practice
(see \S\ref{caveats_visc}). Thus, $\beta<\beta_\mathrm{crit,noniso}$ 
provides an excellent criteria for instability.

\begin{figure}
  \includegraphics[scale=0.4415]{figures/gcorr_compare_vnoniso_maxrate2} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $t_c=\beta\OmK^{-1}$. The vertical dashed line is the critical thermal
    timescale given by Eq. \ref{iso_vsi_cond}. A conjectured critical thermal
    timescale accounting for a non-isothermal background, given by 
     $\beta_\mathrm{crit,noniso}= 0.425$ (see Eq. \ref{bcrit_noniso}),   
    is also shown as the vertical dotted line. 
    \label{bcrit_compare2}}   
\end{figure}

The above calculation demonstrates that the effect of introducing
finite cooling on the VSI operates very similarly in vertically
isothermal disks with an imposed surface, and in vertically
non-isothermal disks with a physical surface. This is despite the fact
that the verticaly buoyancy frequency, $N_z$, formally diverges in
latter case as $|z|\to H_s$.

@@

% We do not find surface modes for $\khat=10$, but do find them for
% sufficiently large $\khat$. We demonstrate this in  % In the vertically
% % non-isothermal disk, surface modes develop for sufficiently large 
% % $\khat$ \citep{barker15}. We demonstrate them in
% Fig. \ref{compare_modes_vnoniso_kx100} which show eigenvalues for
% $\khat=100$. While growth rates generally decrease with increasing 
% $\beta$, we find eigenvalues can overlap, making it difficult to 
% distinguish between different modes. However, since there 
% is a well-defined surface, we may seek the  
% most unstable mode without the boundary-dependent issue that arose 
% with a vertically isothermal disk in the limit $\beta\to0$.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/compare_modes_Gam1.084_kx100.ps}
%   \caption{Same as Fig. \ref{compare_modes_vnoniso_kx10} but with
%     $\khat=100$. In this case surface modes appear nearly parallel to
%     the vertical axis
%     (cf. Fig. \ref{compare_modes_vnoniso_kx10}). However, this
%     distinction becomes difficult for increasing $\beta$ as modes
%     overlap. 
%     % but
%     % overlap with body modes (those that are roughly horizontal when 
%     % $\beta=10^{-3}$) as $\beta$ is further increased. 
%     \label{compare_modes_vnoniso_kx100}}
% \end{figure}

% Fig. \ref{compare_eigenvz_vnoniso_kx100} show the most unstable modes
% with $\khat=100$ and $\beta=10^{-3},\, 10^{-2}$ and
% $0.1$. For $\beta=10^{-3}$ the most 
% unstable mode is an asymmetric surface mode \citepalias[also seen in Fig. 7
% of][]{barker15}. However, increasing the thermal timescale to
% $\beta=10^{-2}$ an anti-symmetric surface mode becomes
% dominant. Further increasing to $\beta=0.1$, the dominant mode has a
% mixed character between a surface mode and the fundamental
% mode. Notice as $\beta$ is increased the perturbation amplitude near
% the vertical boundaries become negligible. This is not surprsing since  
% the stabilizing effect of buoyancy is largest at the disk surface. 
% %formally diverges

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.73cm 0cm
%   0cm]{figures/eigenvectorvz_vnoniso_kx100_beta0d001.ps}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.73cm 0cm
%   0cm]{figures/eigenvectorvz_vnoniso_kx100_beta0d01.ps}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/eigenvectorvz_vnoniso_kx100_beta0d1.ps}
%   \caption{Vertical velocity eigenfunction of the most unstable mode
%     with $\khat=100$ found in the vertically non-isothermal disk with
%     $\beta = 10^{-3}$ (top), $\beta=10^{-2}$ (middle) and $\beta=0.1$
%     (bottom). 
%     \label{compare_eigenvz_vnoniso_kx100}}
% \end{figure}
